---
title: "HSCI 50 LAB 6: Bivariate Analysis for Continuous Variables"
output: pdf_document
header-includes:
- \usepackage[default]{lato}
- \usepackage[T1]{fontenc}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \fancyhead[R]{\ifnum\value{page}>1 \else Haw NJ / ADMU HSP \fi}
mainfont: Lato
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
```

```{r libraries, include = FALSE}
library(tidyverse)
library(knitr)
library(sysfonts)
library(kableExtra)
library(gridExtra)
windowsFonts("Lato" = windowsFont("Lato"))
```

**INSTRUCTIONS**

The data set that we will be using for the entire course is resampled data from the MIT COVID-19 Beliefs, Behaviors & Norms Survey (https://covidsurvey.mit.edu/api.html). This is a multi-country, online survey that examined different COVID-19 perceptions across time, from July 6, 2020 to March 28, 2021. We will be using data from the Philippines aged 20 to 60. **Assume the data are randomly sampled.**

In this lab, you will learn how to conduct and interpret statistical outputs from an independent samples *t*-test, a paired samples *t*-test, and a one-way analysis of variance (ANOVA).

You have many options to submit this worksheet. Either you work on this by hand and scan/take a clear photo of your submission and save as PDF, or type your responses in a Word processor or PowerPoint presentation. You do not have to copy the questions again, but please number them accordingly.

```{r read_data, include = FALSE}
data <- read.csv("C:/Users/neljasonhaw/Google Drive (nhaw@ateneo.edu)/HSCI 50 SY 2021-2022/Labs/vax_norms_dummydata.csv")
```

&nbsp;

Let's revisit the **vaccine norms** variable we had used in Labs 3 and 4. Individuals were asked about their perceptions of how other people will accept the COVID-19 vaccine, as a measurement of vaccine norms. They were asked, "Out of 100 people in your community, how many do you think would take a COVID-19 vaccine if it were made available?" Again, while this is technically a discrete variable, we will assume for the purposes of this class that this is a continuous variable.

&nbsp;

**PART A. Independent samples *t*-test**

For this part, we want to check if there are any differences in vaccine norm by gender (male vs. female) during the last survey wave (March 14 to 28, 2021). The following are statistical outputs that will help you answer the following questions.

```{r eda1, fig.width = 10, fig.height = 5, echo = FALSE, warning = FALSE}
## Preparing data for independent samples t-test
## data frame "data" contains the data used for this lab
## variables: "gender" - male vs female;  "vaccine_norm" - range: 1 to 100; "wave" - 9 (Oct 26 - Nov 8) and 19 (Mar 14 to 28)

# male19 contains all the values of vaccine norms for males during the last survey wave (Wave 19)
male19 <- data %>% filter(gender == "male" & wave == 19) %>% select(vaccine_norm)
male19 <- as.numeric(male19$vaccine_norm)
# female19 contains all the values of vaccine norms for females during the last survey wave (Wave 19)
female19 <- data %>% filter(gender == "female" & wave == 19) %>% select(vaccine_norm)
female19 <- as.numeric(female19$vaccine_norm)

# Filter data - wave 19 only
data19 <- data %>% filter(wave == 19)

# Plot histogram of vaccine norms by gender
hist_gender <- ggplot(data = data19) + 
  geom_histogram(aes(x = vaccine_norm, fill = gender), bins = 20) + 
  theme_minimal() + scale_fill_brewer(palette = "Set2") + 
  facet_wrap(~gender) + xlab("") + ylab("Frequency") + guides(fill = FALSE) + 
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of vaccine norms by gender in the Philippines, March 14 to 28, 2021")
hist_gender

# Boxplot of vaccine norms by gender
boxplot_gender <- ggplot(data = data19) + 
  stat_boxplot(geom = "errorbar", aes(y = vaccine_norm, x = gender)) + 
  geom_boxplot(aes(y = vaccine_norm, x = gender, fill = gender)) + 
  theme_minimal() + scale_fill_brewer(palette = "Set2") + 
  xlab("") + ylab("Vaccine norm") + guides(fill = FALSE) +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Boxplot of vaccine norms by gender in the Philippines, March 14 to 28, 2021")
boxplot_gender
```

&nbsp;

Summary statistics of vaccine norm by gender

```{r summ1, echo = FALSE}
# Summary statistics 
male_centtend <- c(format(round(length(male19), 1), nsmall = 0),
                   format(round(mean(male19), 1), nsmall = 1),
                   format(round(sd(male19), 1), nsmall = 1),
                   format(round(sd(male19)^2, 1), nsmall = 1),
                   round(median(male19), 1),
                   format(round(as.numeric(quantile(male19,0.25)),1), nsmall = 0),
                   format(round(as.numeric(quantile(male19,0.75)),1), nsmall = 0),
                   format(round(as.numeric(quantile(male19,0.75)) - 
                                  as.numeric(quantile(male19,0.25)), 1), nsmall = 0),
                   names(table(male19))[which(table(male19) == max(table(male19)))])
female_centtend <- c(format(round(length(female19), 1), nsmall = 0),
                   format(round(mean(female19), 1), nsmall = 1),
                   format(round(sd(female19), 1), nsmall = 1),
                   format(round(sd(female19)^2, 1), nsmall = 1),
                   round(median(female19), 1),
                   format(round(as.numeric(quantile(female19,0.25)),1), nsmall = 0),
                   format(round(as.numeric(quantile(female19,0.75)),1), nsmall = 0),
                   format(round(as.numeric(quantile(female19,0.75)) - 
                                  as.numeric(quantile(female19,0.25)), 1), nsmall = 0),
                   names(table(female19))[which(table(female19) == max(table(female19)))])
centtend <- as.data.frame(c("Count", "Mean", "Standard deviation (SD)", "Variance", 
                            "Median", "First Quartile (Q1)", "Third Quartile (Q3)", 
                            "Interquartile Range (IQR)", "Mode"))
centtend$male <- male_centtend
centtend$female <- female_centtend
colnames(centtend) <- c("Vaccine norm", "Male", "Female")
kable(centtend, booktabs = "T", align = "lcc")
```

&nbsp;

```{r ttest1}
# male19 contains all the values of vaccine norms for males during the last survey wave (Wave 19)
# female19 contains all the values of vaccine norms for females during the last survey wave (Wave 19)

# Conduct independent samples t-test, two-sided, equal variances assumed
t.test(x = male19, y = female19, alternative = "two.sided", var.equal = TRUE)

# Conduct independent samples t-test, two-sided, unequal variances assumed
t.test(x = male19, y = female19, alternative = "two.sided", var.equal = FALSE)
```


**1. What are the appropriate null and alternative hypothesis for this question?**

**2. What are the assumptions required to conduct an independent samples *t*-test, and how were they fulfilled?**

**3. Which *t*-test did you use - equal or unequal variances assumed? Explain your response.**

**4. What can you conclude? Either use the appropriate test statistic or the 95% confidence interval (CI).**

\newpage


**PART B. Paired samples *t*-test**

In this part, we will compare the vaccine norms between Wave 9 of the survey (Oct 26 to Nov 10, 2020) to the last wave of the survey (Wave 19, Mar 14 to 28, 2021) and see if there was a significant change in vaccine norms over time. The following are statistical outputs that will help you answer the following questions:

&nbsp;

```{r eda2, fig.width = 10, fig.height = 5, echo = FALSE, warning = FALSE}
## Preparing data for paired samples t-test
## data frame "data" contains the data used for this lab
## variables: "gender" - male vs female;  "vaccine_norm" - range: 1 to 100; "wave" - 9 (Oct 26 - Nov 8) and 19 (Mar 14 to 28)
## data already sorted by ID

# wave19 contains all the values of vaccine norms for wave 19
wave19 <- data %>% filter(wave == 19) %>% select(vaccine_norm)
wave19 <- as.numeric(wave19$vaccine_norm)
# wave9 contains all the values of vaccine norms for wave 9
wave9 <- data %>% filter(wave == 9) %>% select(vaccine_norm)
wave9 <- as.numeric(wave9$vaccine_norm)
# generate differences
justwave <- data %>% select(id, wave, vaccine_norm)
justwave_wide <- reshape(justwave, timevar = "wave", direction = "wide")
justwave_wide <- justwave_wide %>% mutate(diff = vaccine_norm.19 - vaccine_norm.9)
diff <- as.numeric(justwave_wide$diff)

# Plot histogram of vaccine norms by wave
hist_wave <- ggplot(data = data) + 
  geom_histogram(aes(x = vaccine_norm, fill = factor(wave)), bins = 20) + 
  theme_minimal() + scale_fill_brewer(palette = "Set1") + 
  facet_wrap(~wave) + xlab("") + ylab("Frequency") + guides(fill = FALSE) + 
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of vaccine norms by survey wave in the Philippines")

hist_wave
```

&nbsp;

```{r eda3, fig.width = 10, fig.height = 5, echo = FALSE}
# Plot difference of vaccine norms
# Histogram
hist_diff <- ggplot(justwave_wide) + geom_histogram(aes(x = diff), bins = 20, fill = "darkmagenta") +
  theme_minimal() + xlab("") + ylab("Frequency") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of differences in vaccine \n norms (Wave 19 - Wave 9)")

# QQ plot
qq_diff <- ggplot(justwave_wide, aes(sample = diff)) + 
  geom_qq(color = "darkmagenta", size = 2) + geom_qq_line(size = 1) +
  theme_minimal() + xlab("Theoretical distribution") + ylab("Sample distribution") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Q-Q plot of differences in vaccine \n norms (Wave 19 - Wave 9)")

grid.arrange(hist_diff, qq_diff, ncol=2)
```

&nbsp;


Summary statistics of vaccine norm by survey wave

```{r summ2, echo = FALSE}
# Summary statistics 
wave9_centtend <- c(format(round(length(wave9), 1), nsmall = 0),
                    format(round(mean(wave9), 1), nsmall = 1),
                   format(round(sd(wave9), 1), nsmall = 1),
                   format(round(sd(wave9)^2, 1), nsmall = 1),
                   round(median(wave9), 1),
                   format(round(as.numeric(quantile(wave9,0.25)),1), nsmall = 0),
                   format(round(as.numeric(quantile(wave9,0.75)),1), nsmall = 0),
                   format(round(as.numeric(quantile(wave9,0.75)) - 
                                  as.numeric(quantile(wave9,0.25)), 1), nsmall = 0),
                   names(table(wave9))[which(table(wave9) == max(table(wave9)))])
wave19_centtend <- c(format(round(length(wave19), 1), nsmall = 0),
                     format(round(mean(wave19), 1), nsmall = 1),
                   format(round(sd(wave19), 1), nsmall = 1),
                   format(round(sd(wave19)^2, 1), nsmall = 1),
                   round(median(wave19), 1),
                   format(round(as.numeric(quantile(wave19,0.25)),1), nsmall = 0),
                   format(round(as.numeric(quantile(wave19,0.75)),1), nsmall = 0),
                   format(round(as.numeric(quantile(wave19,0.75)) - 
                                  as.numeric(quantile(wave19,0.25)), 1), nsmall = 0),
                   names(table(wave19))[which(table(wave19) == max(table(wave19)))])
diff_centtend <- c(format(round(length(diff), 1), nsmall = 0),
                   format(round(mean(diff), 1), nsmall = 1),
                   format(round(sd(diff), 1), nsmall = 1),
                   format(round(sd(diff)^2, 1), nsmall = 1),
                   round(median(diff), 1),
                   format(round(as.numeric(quantile(diff,0.25)),1), nsmall = 0),
                   format(round(as.numeric(quantile(diff,0.75)),1), nsmall = 0),
                   format(round(as.numeric(quantile(diff,0.75)) - 
                                  as.numeric(quantile(diff,0.25)), 1), nsmall = 0),
                   names(table(diff))[which(table(diff) == max(table(diff)))])

centtend2 <- as.data.frame(c("Count", "Mean", "Standard deviation (SD)", "Variance", 
                            "Median", "First Quartile (Q1)", "Third Quartile (Q3)", 
                            "Interquartile Range (IQR)", "Mode"))
centtend2$wave9 <- wave9_centtend
centtend2$wave19 <- wave19_centtend
centtend2$diff <- diff_centtend
colnames(centtend2) <- c("Vaccine norm", "Wave 9", "Wave 19", "Difference (Wave 19 - Wave 9)")
kable(centtend2, booktabs = "T", align = "lccc")
```

&nbsp;

```{r ttest2}
# wave9 contains all the values of vaccine norms for wave 9
# wave19 contains all the values of vaccine norms for wave 19
# data already sorted to align on the same row for the same ID number of survey participant

# Conduct paired samples t-test, two-sided
t.test(x = wave9, y = wave19, alternative = "two.sided", paired = TRUE)
```

**1. What are the appropriate null and alternative hypothesis for this question?**

**2. What are the assumptions required to conduct a paired samples *t*-test, and how were they fulfilled?**

**3. What can you conclude? Either use the appropriate test statistic or the 95% confidence interval (CI).**

\newpage


**PART C. One-way analysis of variance**


For this part, we want to check if there are any differences in vaccine norm by age group during the last survey wave (March 14 to 28, 2021). The following are statistical outputs that will help you answer the following questions.

```{r eda4, fig.width = 10, fig.height = 8, echo = FALSE, warning = FALSE}
## Preparing data for one way ANOVA
## data frame "data" contains the data used for this lab
## variables: "gender" - male vs female;  "vaccine_norm" - range: 1 to 100; "wave" - 9 (Oct 26 - Nov 8) and 19 (Mar 14 to 28)


# Plot histogram of vaccine norms by age group
hist_agegrp <- ggplot(data = data19) + 
  geom_histogram(aes(x = vaccine_norm, fill = age_grp), bins = 20) + 
  theme_minimal() + scale_fill_brewer(palette = "Set1") + 
  facet_wrap(~age_grp) + xlab("") + ylab("Frequency") + guides(fill = FALSE) + 
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of vaccine norms by age group in the Philippines, March 14 - 28, 2021")

hist_agegrp
```

&nbsp;

Summary statistics of vaccine norm by age group

```{r summ3, echo = FALSE, message = FALSE, warning = FALSE}
# Summary statistics 
agetable <- data19 %>% group_by(age_grp) %>% summarize(
  Count = n(),
  Mean = format(round(mean(vaccine_norm), 1), nsmall = 1),
  SD = format(round(sd(vaccine_norm), 1), nsmall = 1),
  Variance = format(round(sd(vaccine_norm)^2, 1), nsmall = 1),
  Median = round(median(vaccine_norm), 1),
  Q1 = format(round(as.numeric(quantile(vaccine_norm,0.25)),1), nsmall = 1),
  Q3 = format(round(as.numeric(quantile(vaccine_norm,0.75)),1), nsmall = 1),
  IQR = format(round(as.numeric(quantile(vaccine_norm,0.75)) - as.numeric(quantile(vaccine_norm,0.25)), 1), nsmall = 1)) %>%
  gather(key = key, value = value, -age_grp) %>% 
  spread(key = age_grp, value = value)

agetable <- agetable %>% arrange(match(key, c("Count", "Mean", "SD", "Variance", "Median", "Q1", "Q3", "IQR")))
colnames(agetable) <- c("Vaccine norm", "20-30 y/o", "31-40 y/o", "41-50 y/o", "51-60 y/o")
kable(agetable, booktabs = "T", align = "lcccc")
```

&nbsp;

```{r anova}
# The "vaccine_norm" variable contains the vaccine norm scores
# The "age_grp" variable contains the age group categories

# Conduct one-way ANOVA
summary(aov(vaccine_norm ~ age_grp, data = data19))

# Conduct post hoc pairwise t-test with Bonferroni correction
# The results shown are the Bonferroni-inflated p-values, meaning the p-values have been inflated
# with the Bonferroni correction factor already. Compare with the original alpha.
pairwise.t.test(data19$vaccine_norm, data19$age_grp, p.adj = "bonf")
```


**1. What are the appropriate null and alternative hypothesis for this question?**

**2. What are the assumptions required to conduct a one-way analysis of variance (ANOVA), and how were they fulfilled?**

**3. What can you conclude from the one-way ANOVA AND the post-hoc pairwise *t*-test with Bonferroni correction (if applicable)? Mention the relevant test statistic and *p*-values, as well as the mean values of vaccine norm in each group**

\center END OF LAB \center