---
title: "HSCI 50 LAB 7: Correlation and Regression"
output: pdf_document
header-includes:
- \usepackage[default]{lato}
- \usepackage[T1]{fontenc}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \fancyhead[R]{\ifnum\value{page}>1 \else Haw NJ / ADMU HSP \fi}
mainfont: Lato
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
```

```{r libraries, include = FALSE}
library(tidyverse)
library(knitr)
library(sysfonts)
library(kableExtra)
library(gridExtra)
windowsFonts("Lato" = windowsFont("Lato"))
```

**INSTRUCTIONS**

The data set that we will be using for the entire course is resampled data from the MIT COVID-19 Beliefs, Behaviors & Norms Survey (https://covidsurvey.mit.edu/api.html). This is a multi-country, online survey that examined different COVID-19 perceptions across time, from July 6, 2020 to March 28, 2021. We will be using data from the Philippines aged 20 to 60. **Assume the data are randomly sampled.**

In this lab, you will learn how to interpret correlation outputs between two continuous variables, and interpret outputs from simple and multiple linear and logistic regression models. 

You have many options to submit this worksheet. Either you work on this by hand and scan/take a clear photo of your submission and save as PDF, or type your responses in a Word processor or PowerPoint presentation. You do not have to copy the questions again, but please number them accordingly.

```{r read_data, include = FALSE}
data_norms <- read.csv("C:/Users/neljasonhaw/Google Drive (nhaw@ateneo.edu)/HSCI 50 SY 2021-2022/Labs/vax_norms_dummydata.csv")
data_accept <- read.csv("C:/Users/neljasonhaw/Google Drive (nhaw@ateneo.edu)/HSCI 50 SY 2021-2022/Labs/vax_accept_dummydata.csv")
```

&nbsp;

**PART A. Correlation**

Let us revisit the **vaccine norms** variable we had used in Labs 3, 4 and 6. Individuals were asked about their perceptions of how other people will accept the COVID-19 vaccine, as a measurement of vaccine norms. They were asked, "Out of 100 people in your community, how many do you think would take a COVID-19 vaccine if it were made available?" Again, while this is technically a discrete variable, we will assume for the purposes of this class that this is a continuous variable.

In Lab 6, we determined that vaccine norms differed by age group. We will examine the same relationship but with a different functional form of the age variable. We will use age as a continuous variable in years. While this is technically a discrete variable, we will assume for the purposes of this class that this is continuous.

The following are statistical outputs that will help you answer the following questions.

&nbsp;

```{r histcorr, fig.height = 5, fig.width = 10, echo = FALSE, message = FALSE}
# Filter data to only include those from Wave 19
data_norms_19 <- data_norms %>% filter(wave == 19)

# Histogram of age
hist_age <- ggplot(data = data_norms_19) + 
  geom_histogram(aes(x = age), bins = 20, fill = "darkmagenta") +
  theme_minimal() + xlab("") + ylab("Frequency") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of age")

# Histogram of vaccine_norm
hist_vax <- ggplot(data = data_norms_19) + 
  geom_histogram(aes(x = vaccine_norm), bins = 20, fill = "darkorange2") +
  theme_minimal() + xlab("") + ylab("Frequency") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Histogram of vaccine norm")

# Arrange histograms side by side
grid.arrange(hist_age, hist_vax, ncol=2)

# Scatterplot of vaccine_norm by age
scatter1 <- ggplot(data = data_norms_19, aes(x = age, y = vaccine_norm)) + 
  geom_point(color = "gray") + 
  geom_smooth(method = lm, se = FALSE, color = "deepskyblue4", size = 2) +
  theme_minimal() + xlab("Age (in years)") + ylab("Vaccine norm") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Scatterplot of vaccine norm by age in the Philippines, March 14 to 28, 2021")
scatter1
```

&nbsp;

```{r corrtest, warning = FALSE, message = FALSE}
# Pearson's correlation coefficient output for vaccine norm by age relationship
cor.test(x = data_norms_19$age, y = data_norms_19$vaccine_norm, method = 'pearson')

# Spearman's correlation coefficient output for vaccine norm by age relationship
cor.test(x = data_norms_19$age, y = data_norms_19$vaccine_norm, method = 'spearman')
```

&nbsp;

&nbsp;

**1. What are the appropriate null and alternative hypotheses for the correlation above?**


**2. What is the scatterplot telling us? Is there some indication of a potentially linear relationship?**

**3. Based on the scatterplot, is it appropriate to calculate a Pearson's correlation coefficient, a Spearman's correlation coefficient, both, or neither?**

**4. What are the assumptions required for calculating the correlation coefficient?**


**5. What can you conclude? Either use the *p*-value or the 95% confidence interval (CI). Report the magnitude and direction of the correlation coefficient. Report Pearson's/Spearman's correlation coefficients as appropriate.**


&nbsp;


**PART B. Linear Regression**

We will now try to model the relationship of vaccine norm against age and gender. Gender is a binary variable in our dataset (Male / Female) while we will model age as a continuous variable centered at 20 years old, which is the lowest age in our data. Note that centering is not restricted to the mean of the data, we can center the value of any continuous variable to any constant value as approrpiate to make the model more interpretable.

The following are statistical outputs that will help you answer the following questions. Aside from these statistical outputs, consider the statistical outputs from the previous question. Recall that centering a variable does not change the shape of the scatterplot, so the same scatterplot of vaccine norm by age applies to this part regardless of where we centered the variable.

When reading the statistical outputs of the linear regression model, pay attention to the `Analysis of Variance` table and the table of `Coefficients`. Under the table of `Coefficients`:

* `(Intercept)` shows the value of $b_0$
* The name of the variable shows the value of $b_1$
* The `t-value` is the t-statistic
* `Pr(>|t|)` shows the two-sided *p*-value of the t-statistic for the coefficient
* There are also 95% confidence intervals of the beta coefficients indicated under `2.5%` for the lower limit of the confidence interval and `97.5%` for the upper limit of the confidence interval.

The coefficient of determination is indicated under `Multiple R-squared`

There are also a set of diagnostic plots that should look familiar to you: a residuals vs. fitted values scatterplot, and a Q-Q plot of the residuals.

&nbsp;

**1. First, take a look at the statistical output of the simple linear regression of vaccine norm by age (Statistical Output 1). What are the appropriate null and alternative hypotheses for that particular linear regression model?**

**2. What are the assumptions required for the simple linear regression of vaccine norm by age (Statistical Output 1)? Do we expect the same set of assumptions for the simple linear regression of vaccine norm by gender?**

**3. Describe the equation of the line for all three regression models above, with the values of the coefficients from the regression model results. Recall that the format of the equation of the line for a simple linear regression is $Y = \beta_0 + \beta_1X_1$ and add more $\beta_pX_p$ as needed in a multiple linear regression model.**

**4. For each of the three regression models above, what do they altogether say about the association of vaccine norm by age and gender? Either use the appropriate test statistic or the 95% confidence interval (CI). Report an interpretation for the point estimate of age in the simple linear regression model. Report an interpretation for the point estimate of age in the multiple linear regression as well.**

**5. Using the appropriate equation of the line from 3, predict the vaccine norm score for a 50 year old male. Report the point estimate only.**

\newpage

**Statistical Outputs 1: Simple Linear Regression of Vaccine Norms by Age (centered at  20)**

```{r linreg_age, fig.height = 5, fig.width = 10,  echo = FALSE, message = FALSE}
# Center age at age 20 and call the variable age_cent20
data_norms_19 <- data_norms_19 %>% mutate(age_cent20 = age - 20)

## Simple linear regression of vaccine norm by age (centered at 20)
# Model
simple_age <- lm(vaccine_norm  ~ age_cent20, data = data_norms_19)
# ANOVA of regression model
anova(simple_age)
# Model coefficients
summary(simple_age)
confint(simple_age)
# Predict residuals
data_norms_19$age_residuals <- residuals(simple_age)
# Predict fitted values
data_norms_19$age_predict <- predict(simple_age)
# Scatterplot of residuals vs. fitted values
scatter2 <- ggplot(data = data_norms_19, aes(x = age_predict, y = age_residuals)) + 
  geom_point(color = "gray") + 
  geom_smooth(method = lm, se = FALSE, color = "deepskyblue4", size = 2) +
  theme_minimal() + xlab("Fitted values") + ylab("Residuals") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Residuals vs. Fitted values scatterplot")
# Q-Q plot of residuals
qq2 <- ggplot(data = data_norms_19, aes(sample = age_residuals)) + 
  geom_qq(color = "deepskyblue4", size = 2) + geom_qq_line(size = 1) +
  theme_minimal() + xlab("Theoretical distribution") + ylab("Sample distribution") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Q-Q plot of residuals")
# Arrange plots side by side
grid.arrange(scatter2, qq2, ncol=2)
```

\newpage

**Statistical Outputs 2: Simple Linear Regression of Vaccine Norms by Gender**

For gender, the reference category is female.

```{r linreg_gender, fig.height = 5, fig.width = 10,  echo = FALSE, message = FALSE}
## Simple linear regression of vaccine norm by age (centered at 20)
# Model
simple_gender <- lm(vaccine_norm  ~ gender, data = data_norms_19)
# ANOVA of regression model
anova(simple_gender)
# Model coefficients
summary(simple_gender)
confint(simple_gender)
# Predict residuals
data_norms_19$gender_residuals <- residuals(simple_gender)
# Predict fitted values
data_norms_19$gender_predict <- predict(simple_gender)
# Scatterplot of residuals vs. fitted values
scatter3 <- ggplot(data = data_norms_19, aes(x = gender_predict, y = gender_residuals)) + 
  geom_point(color = "gray") + 
  geom_smooth(method = lm, se = FALSE, color = "deepskyblue4", size = 2) +
  theme_minimal() + xlab("Fitted values") + ylab("Residuals") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Residuals vs. Fitted values scatterplot")
# Q-Q plot of residuals
qq3 <- ggplot(data = data_norms_19, aes(sample = gender_residuals)) + 
  geom_qq(color = "deepskyblue4", size = 2) + geom_qq_line(size = 1) +
  theme_minimal() + xlab("Theoretical distribution") + ylab("Sample distribution") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Q-Q plot of residuals")
# Arrange plots side by side
grid.arrange(scatter3, qq3, ncol=2)
```

\newpage

**Statistical Outputs 3: Multiple Linear Regression of Vaccine Norms by Gender and Age (centered at  20)**

For gender, the reference category is female.

```{r linreg_multiple, fig.height = 4, fig.width = 10,  echo = FALSE, message = FALSE}
## Multiple linear regression of vaccine norm by gender and age (centered at 20)
# Model
multiple_line <- lm(vaccine_norm  ~ gender + age_cent20, data = data_norms_19)
# ANOVA of regression model
anova(multiple_line)
# Model coefficients
summary(multiple_line)
confint(multiple_line)
# Predict residuals
data_norms_19$multiple_line_residuals <- residuals(multiple_line)
# Predict fitted values
data_norms_19$multiple_line_predict <- predict(multiple_line)
# Scatterplot of residuals vs. fitted values
scatter4 <- ggplot(data = data_norms_19, aes(x = multiple_line_predict, y = multiple_line_residuals)) + 
  geom_point(color = "gray") + 
  geom_smooth(method = lm, se = FALSE, color = "deepskyblue4", size = 2) +
  theme_minimal() + xlab("Fitted values") + ylab("Residuals") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Residuals vs. Fitted values scatterplot")
# Q-Q plot of residuals
qq4 <- ggplot(data = data_norms_19, aes(sample = multiple_line_residuals)) + 
  geom_qq(color = "deepskyblue4", size = 2) + geom_qq_line(size = 1) +
  theme_minimal() + xlab("Theoretical distribution") + ylab("Sample distribution") +
  theme(text=element_text(family="Lato", size = 16), panel.grid = element_blank()) +
  ggtitle("Q-Q plot of residuals")
# Arrange plots side by side
grid.arrange(scatter4, qq4, ncol=2)
```



\newpage


**PART C. Logistic regression**

For this part, we will recode the vaccine acceptance variable into a binary category similar to Lab 3 and 5 called vaccine hesitancy, where those who do not accept the vaccine or are unsure about it will be considered vaccine hesitant, otherwise they are not vaccine hesitant. Age is again analyzed as a continuous variable centered at age 20. 

The following are statistical outputs that will help you answer the following questions. 

When reading the statistical outputs of the logistic regression model, pay attention to the following:

* The `Coefficients` table is like the `Coefficients` table in the linear regression model result; however, the coefficients have not yet been exponentiated to odds ratios.
* The odds ratios and corresponding 95% CI are found at the end of the output under the table with the header columns `2.5%` and `97.5%`

&nbsp;

**Statistical Output 4: Simple Logistic Regression of Vaccine Hesitancy by Age (centered at 20)**

```{r logistic1, fig.height = 4, fig.width = 10,  echo = FALSE, message = FALSE}
# Filter data to only include those from Wave 19
data_accept_19 <- data_accept %>% filter(wave == 19) 

# Generate binary variable of vaccine hesitancy
data_accept_19 <- data_accept_19 %>% 
  mutate(hesitant = case_when(response == "Yes" | response == "Vaccinated" ~ 0,
                              response == "No" | response == "Don't Know" ~ 1))

# Center age at age 20 and call the variable age_cent20
data_accept_19 <- data_accept_19 %>% mutate(age_cent20 = age - 20)

## Simple linear regression of vaccine norm by age (centered at 20)
# Model
age_logit <- glm(hesitant ~ age_cent20, family = binomial(link = 'logit'), data = data_accept_19)
# Model coefficients
summary(age_logit)
exp(cbind(coef(age_logit), confint(age_logit)))  
```

\newpage

**Statistical Output 5: Simple Logistic Regression of Vaccine Hesitancy by Gender**

For gender, the reference category is female.

```{r logistic2, fig.height = 4, fig.width = 10,  echo = FALSE, message = FALSE}
## Simple linear regression of vaccine norm by gender
# Model
age_logit <- glm(hesitant ~ gender, family = binomial(link = 'logit'), data = data_accept_19)
# Model coefficients
summary(age_logit)
exp(cbind(coef(age_logit), confint(age_logit)))  
```

\newpage

**Statistical Output 6: Multiple Logistic Regression of Vaccine Hesitancy by Gender and Age (centered at  20)**

For gender, the reference category is female.

```{r logistic3, fig.height = 4, fig.width = 10,  echo = FALSE, message = FALSE}
## Multiple linear regression of vaccine norm by gender and age (centered at 20)
# Model
multiple_logit <- glm(hesitant  ~ gender + age_cent20, family = binomial(link='logit'), data = data_accept_19)
# Model coefficients
summary(multiple_logit)
exp(cbind(coef(multiple_logit), confint(multiple_logit)))  
```

&nbsp;

**For each of the three logistic regression models above, what do they altogether say about the association of vaccine hesitancy by age and gender? Either use the appropriate test statistic or the 95% confidence interval (CI). Provide an interpretation of the odds ratio between male and female from the simple logistic regression model and the multiple logistic regression model.**


\center END OF LAB \center

